---
title: "Assignment 1"
author: "Justine Lang"
date: "2023-03-27"
output: 
  html_document: 
    code_folding: hide
---

```{r setup, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(janitor)
library(vegan)
library(tidyr)
library(dplyr)
library(kableExtra)
```


```{r}
#read presence-absence data

grasspa <- read.csv(here("data", "grassland_dat_pa.csv"),header=TRUE)

padat <- grasspa[,4:40]	#plot x species P/A matrix in format for vegan analysis
```

```{r}
#read cover data
	grasscov <- read.csv(here("data", "grassland_dat_cover.csv"),header=TRUE)
	covdat <- grasscov[,4:40]	#plot x species cover matrix in format for vegan analysis
```

```{r}
### read in 2023 data

grasspa_2023 <- read_csv(here("data", "grassland_dat_pa_2023.csv"))
padat_2023 <- grasspa_2023[,4:27]

grasscov_2023 <- read_csv(here("data", "grassland_dat_cover_2023.csv"))
covdat_2023 <- grasscov_2023[,4:27]

```


# 1. Species Richness

## 1a)

### Accumulation Curves - 10 Subplots (2019 & 2023)


```{r}
#create exact species accumulation curves for 4 plots using 10 subplots (2019)
	accum1 <- specaccum(padat[1:10,]) 	#plot 1, subplots 1-10
	accum2 <- specaccum(padat[13:22,])	#plot 2, subplots 1-10
	accum3 <- specaccum(padat[25:34,])	#plot 3, subplots 1-10
	accum4 <- specaccum(padat[37:46,])	#plot 4, subplots 1-10
```

```{r}
#create exact species accumulation curves for 3 plots using 10 subplots (2023)
	accum1_23 <- specaccum(padat_2023[1:10,]) 	#plot 1, subplots 1-10
	accum2_23 <- specaccum(padat_2023[13:22,])	#plot 2, subplots 1-10
	accum3_23 <- specaccum(padat_2023[24:33,])	#plot 3, subplots 1-10
```

```{r figure 1, fig.align='center', fig.cap= "Figure 1: Plant species accumulation curves for 2019 and 2023 utilizing 10 subplots per plot. 4 plots were sampled in 2019. 3 plots were sampled in 2023. Plots were located in the valley (V) and on the north (N and N2) and south (S) slopes."}
#plot accumulation curves for each plot
	plot(accum1,ylim=c(0,20),xlab="subplots",ylab="species")
		lines(accum2,col="red")
		lines(accum3,col="blue")
		lines(accum4,col="green")
		lines(accum1_23, col = "orange")
		lines(accum2_23, col = "skyblue")
		lines(accum3_23, col = "yellow")
	 	text(10.2,9,"V")	#valley plot
	 	text(10.2,14,"N") #North plot
	 	text(10.2,16,"S") #South plot
	 	text(10.2,18,"N2") #second north plot
	 	text(10.2, 10, "V23") ### V 2023 plot
	 	text(10.2, 11, "N23") ### N 2023 plot
	 	text(10.2, 15, "S23") ### S 2023 plot
```


### Accumulation Curves - 12 Subplots (2019 & 2023)

```{r}
#create exact species accumulation curves for 4 plots using 12 subplots (2019)
	accum1b <- specaccum(padat[1:12,]) 	#plot 1, subplots 1-10
	accum2b <- specaccum(padat[13:24,])	#plot 2, subplots 1-10
	accum3b <- specaccum(padat[25:36,])	#plot 3, subplots 1-10
	accum4b <- specaccum(padat[37:48,])	#plot 4, subplots 1-10
```

```{r}
#create exact species accumulation curves for 3 plots using 11-12 subplots (2023)
	accum1b_23 <- specaccum(padat_2023[1:12,]) 	#plot 1, subplots 1-10
	accum2b_23 <- specaccum(padat_2023[13:23,])	#plot 2, subplots 1-10
	accum3b_23 <- specaccum(padat_2023[24:34,])	#plot 3, subplots 1-10
```

```{r figure 2, fig.align='center', fig.cap="Figure 2: Plant species accumulation curves for 2019 and 2023 utilizing 12 subplots per plot. 4 plots were sampled in 2019. 3 plots were sampled in 2023. Plots were located in the valley (V) and on the north (N and N2) and south (S) slopes."}
#plot accumulation curves for each plot, with subplots 11 and 12
	plot(accum1b,ylim=c(0,25),xlab="subplots",ylab="species")
		lines(accum2b,col="red")
		lines(accum3b,col="blue")
		lines(accum4b,col="green")
		lines(accum1b_23, col = "orange")
		lines(accum2b_23, col = "skyblue")
		lines(accum3b_23, col = "yellow")
	 	text(12.1,11,"V")	#valley plot
	 	text(12.1,15,"N") #North plot
	 	text(12.1,17,"S") #South plot
	 	text(12.1,23,"N2") #second north plot
	 	text(12.1, 12, "V23") ### valley 2023 plot
	 	text(11.1, 12.5, "N23") ### north 2023 plot
	 	text(11.1, 16, "S23") ### south 2023 plot
```


## 1b)

### Single Curve - 40 Subplots

```{r}
#plot single accumulation curve using 40 (2019) and 30 (2023) subplots
accum_allsubplots <- specaccum(padat[c(1:10,13:22,25:34,37:46),])
accum_allsubplots_23 <- specaccum(padat_2023[c(1:10, 13:22, 24:33),])

	plot(accum_allsubplots,xlab="subplots",ylab="species")
	lines(accum_allsubplots_23, col="blue")
	text(40.2, 34.5, "19")
	text(30.3, 23, "23")
```



### Single Curve - 48 Subplots

```{r}
#plot single accumulation curve using 48 (2019) and 34 (2023) subplots
accum_allsubplotsb <- specaccum(padat[c(1:12,13:24,25:36,37:48),])
accum_allsubplotsb_23 <- specaccum(padat_2023[c(1:12, 13:23, 24:34),])

	plot(accum_allsubplotsb,xlab="subplots",ylab="species")
	lines(accum_allsubplotsb_23, col="blue")
	text(48, 37, "19")
	text(35, 25, "23")
```


### Chao Estimate

```{r}
#calculate Bray-Curtis dissimilarity between plots
#First, aggregate subplot data
	p1 <- apply(padat[1:12,],2,max)
	p2 <- apply(padat[13:24,],2,max)
	p3 <- apply(padat[25:36,],2,max)
	p4 <- apply(padat[37:48,],2,max)
# create new matrix to calculate pairwise plot dissimilarity
	plots_spp <- rbind(p1,p2,p3,p4)
#produce dissimilarity matrix
	plots_bray_diss <- vegdist(plots_spp,method="bray")
```

**2019**

```{r}
specpool(plots_spp)
```

**2023**

```{r}
p1_23 <- apply(padat_2023[1:12,],2,max)
p2_23 <- apply(padat_2023[13:23,],2,max)
p3_23 <- apply(padat_2023[24:34,],2,max)

plots_spp_23 <- rbind(p1_23,p2_23,p3_23)
```

```{r}
specpool(plots_spp_23)
```

## 1c)

### Native

```{r}
### read in species data

species <- read_csv(here("data", "grassland_common_species_ex1.csv"))

# subplots_40 <- padat[c(1:10,13:22,25:34,37:46),]

# accum_all <- specaccum(subplots_40)

# plot_sub <- plot(accum_all, xlab = "subplots", ylab = "species")

native <- species %>%
  filter(Native == 1) %>% 
  filter(is.na(Native) == F)

exotic <- species %>%
  filter(Native == 0) %>% 
  filter(is.na(Native) == F)

native_wide <- native %>% 
  pivot_wider(names_from = Code,
              values_from = Native)

sub_native <- padat %>% 
  select(any_of(names(native_wide)))

sub_native_23 <- padat_2023 %>% 
  select(any_of(names(native_wide)))

```

```{r}
#plot single accumulation curve using 40 subplots for native species

subnative_list <- sub_native[c(1:10,13:22,25:34,37:46),] %>% 
  drop_na()

accum_allnative <- specaccum(subnative_list)
accum_allnative_23 <- specaccum(sub_native_23[c(1:10, 13:22, 24:33),])

plot(accum_allnative, xlab="subplots",ylab="species")
lines(accum_allnative_23, col="green4")
text(40.1, 26, "19")
text(30.2, 13, "23")

```

**2019**

```{r}
specpool(sub_native)
```

**2023**

```{r}
specpool(sub_native_23)
```


## 1d)

### Exotic

```{r}
exotic_wide <- exotic %>% 
  pivot_wider(names_from = Code,
              values_from = Native)

sub_exotic <- padat %>% 
  select(any_of(names(exotic_wide)))

sub_exotic_23 <- padat_2023 %>% 
  select(any_of(names(exotic_wide)))
```

```{r}
#plot single accumulation curve using 40 subplots for exotic species

subexotic_list <- sub_exotic[c(1:10,13:22,25:34,37:46),] %>% 
  drop_na()

accum_allexotic <- specaccum(subexotic_list)
accum_allexotic_23 <- specaccum(sub_exotic_23[c(1:10, 13:22, 24:33),])

plot(accum_allexotic, xlab="subplots",ylab="species")
lines(accum_allexotic_23, col="red2")
text(40.1, 9, "19")
text(32, 5, "23")
```

**2019**

```{r}
specpool(sub_exotic)
```

**2023**

```{r}
specpool(sub_exotic_23)
```

**How do accumulation curves and estimated sizes of species pools compare for native compared to exotic species?**

Answer here. 


# 2. Community Dissimilarity

## 2a)

### Bray-Curtis

**2019**

```{r}
plots_bray_diss
```


**2023**

```{r}
plots_bray_diss_23 <- vegdist(plots_spp_23,method="bray")

plots_bray_diss_23
```

## 2c)

### Native

**2019**

```{r}
# sub_native_clean <- sub_native

#calculate Bray-Curtis dissimilarity between plots
#First, aggregate subplot data
	p1n <- apply(sub_native[1:10,],2,max)
	p2n <- apply(sub_native[13:22,],2,max)
	p3n <- apply(sub_native[25:34,],2,max)
	p4n <- apply(sub_native[37:46,],2,max)
# create new matrix to calculate pairwise plot dissimilarity
	plots_spp_n <- rbind(p1n,p2n,p3n,p4n)
#produce dissimilarity matrix
plots_bray_diss_n <- vegdist(plots_spp_n,method="bray")
```

```{r}
plots_bray_diss_n
```

**2023**

```{r}
#calculate Bray-Curtis dissimilarity between plots
#First, aggregate subplot data
	p1n_23 <- apply(sub_native_23[1:10,],2,max)
	p2n_23 <- apply(sub_native_23[13:22,],2,max)
	p3n_23 <- apply(sub_native_23[24:33,],2,max)

# create new matrix to calculate pairwise plot dissimilarity
	plots_spp_n_23 <- rbind(p1n_23,p2n_23,p3n_23)
#produce dissimilarity matrix
plots_bray_diss_n_23 <- vegdist(plots_spp_n_23,method="bray")
```

```{r}
plots_bray_diss_n_23
```

## 2b)

### Cover Data

**2019**

```{r}
#Analyze cover data
#Calculate area-weighted cover, total area sampled = 10 *0.25 + 2*2.5 = 7.5 sq m
	p1_cov <- (2.5/7.5)*apply(covdat[1:10,],2,mean) + 
			(5/7.5)*apply(covdat[11:12,],2,mean)
	p2_cov <- (2.5/7.5)*apply(covdat[13:22,],2,mean) + 
			(5/7.5)*apply(covdat[23:24,],2,mean)
	p3_cov <- (2.5/7.5)*apply(covdat[25:34,],2,mean) + 
			(5/7.5)*apply(covdat[35:36,],2,mean)
	p4_cov <- (2.5/7.5)*apply(covdat[37:46,],2,mean) + 
			(5/7.5)*apply(covdat[47:48,],2,mean)
```

```{r}
#Analyze plot dissimilarity based on cover
	plots_cov <- rbind(p1_cov,p2_cov,p3_cov,p4_cov)
	plots_euclid_diss <- vegdist(plots_cov,method="euclidean")
	plots_hclust <- hclust(plots_euclid_diss,method="average")		#cluster analysis of plots (contrived)
		plot(plots_hclust)
	plots_hclust_pa <- hclust(plots_bray_diss,method="average")
		plot(plots_hclust_pa)
```

**2023**

```{r}
#Analyze cover data
#Calculate area-weighted cover, total area sampled = 10 *0.25 + 2*2.5 = 7.5 sq m
	p1_cov_23 <- (2.5/7.5)*apply(covdat_2023[1:10,],2,mean) + 
			(5/7.5)*apply(covdat_2023[11:12,],2,mean)
	p2_cov_23 <- (2.5/7.5)*apply(covdat_2023[13:22,],2,mean) + 
			(2.5/5)*apply(covdat_2023[23,],2,mean)
	p3_cov_23 <- (2.5/7.5)*apply(covdat_2023[24:33,],2,mean) + 
			(2.5/5)*apply(covdat_2023[34,],2,mean)
```

```{r}
#Analyze plot dissimilarity based on cover
	plots_cov_23 <- rbind(p1_cov_23,p2_cov_23,p3_cov_23)
	plots_euclid_diss_23 <- vegdist(plots_cov_23,method="euclidean")
	plots_hclust_23 <- hclust(plots_euclid_diss_23,method="average")		#cluster analysis of plots (contrived)
		plot(plots_hclust_23)
	plots_hclust_pa_23 <- hclust(plots_bray_diss_23,method="average")
		plot(plots_hclust_pa_23)
```

## 2d)

### Native

**2019**

```{r}
### Filter by native

sub_native_cover <- covdat %>% 
  select(any_of(names(native_wide)))
```

```{r}
#Analyze cover data
#Calculate area-weighted cover, total area sampled = 10 *0.25 + 2*2.5 = 7.5 sq m
	p1_covn <- (2.5/7.5)*apply(sub_native_cover[1:10,],2,mean) + 
			(5/7.5)*apply(sub_native_cover[11:12,],2,mean)
	p2_covn <- (2.5/7.5)*apply(sub_native_cover[13:22,],2,mean) + 
			(5/7.5)*apply(sub_native_cover[23:24,],2,mean)
	p3_covn <- (2.5/7.5)*apply(sub_native_cover[25:34,],2,mean) + 
			(5/7.5)*apply(sub_native_cover[35:36,],2,mean)
	p4_covn <- (2.5/7.5)*apply(sub_native_cover[37:46,],2,mean) + 
			(5/7.5)*apply(sub_native_cover[47:48,],2,mean)
```

```{r}
#Analyze plot dissimilarity based on cover
	plots_cov_n <- rbind(p1_covn,p2_covn,p3_covn,p4_covn)
	plots_euclid_diss_n <- vegdist(plots_cov_n,method="euclidean")
	plots_hclust_n <- hclust(plots_euclid_diss_n,method="average")		#cluster analysis of plots (contrived)
		plot(plots_hclust_n)
	plots_hclust_pa_n <- hclust(plots_bray_diss_n,method="average")
		plot(plots_hclust_pa_n)
```


**2023**

```{r}
sub_native_cov_23 <- covdat_2023 %>% 
  select(any_of(names(native_wide)))
```

```{r}
	p1_covn_23 <- (2.5/7.5)*apply(sub_native_cov_23[1:10,],2,mean) + 
			(5/7.5)*apply(sub_native_cov_23[11:12,],2,mean)
	p2_covn_23 <- (2.5/7.5)*apply(sub_native_cov_23[13:22,],2,mean) + 
			(2.5/5)*apply(sub_native_cov_23[23,],2,mean)
	p3_covn_23 <- (2.5/7.5)*apply(sub_native_cov_23[24:33,],2,mean) + 
			(2.5/5)*apply(sub_native_cov_23[34,],2,mean)
```

```{r}
#Analyze plot dissimilarity based on cover
	plots_covn_23 <- rbind(p1_covn_23,p2_covn_23,p3_covn_23)
	plots_euclid_diss_n23 <- vegdist(plots_covn_23,method="euclidean")
	plots_hclust_n23 <- hclust(plots_euclid_diss_n23,method="average")		#cluster analysis of plots (contrived)
		plot(plots_hclust_n23)
	plots_hclust_pa_n23 <- hclust(plots_bray_diss_n_23,method="average")
		plot(plots_hclust_pa_n23)
```


**Briefly discuss the patterns of plot dissimilarity based on P/A compared to cover data. How do they differ and why? Next, briefly discuss patterns of plot dissimilarity based on all species vs. native species only.**

Answer here.


# 3. Community Evenness

## 3a & 3b)

### Plot Diversity

**2019**

```{r}
#Analyze plot diversity
	plots_shannon <- diversity(plots_cov,index="shannon") 	#shannon H'
	plots_simpson <- diversity(plots_cov,index="simpson")	#simpson diversity
	plots_richness <- specnumber(plots_cov)					#species richness
	
	plots_shannon
	plots_simpson
	plots_richness
```

**2023**

```{r}
#Analyze plot diversity
	plots_shannon_23 <- diversity(plots_cov_23,index="shannon") 	#shannon H'
	plots_simpson_23 <- diversity(plots_cov_23,index="simpson")	#simpson diversity
	plots_richness_23 <- specnumber(plots_cov_23)					#species richness
	
	plots_shannon_23
	plots_simpson_23
	plots_richness_23
```

## 3c & 3d)

### Native

**2019**

```{r}
plots_sh_n <- diversity(plots_cov_n,index="shannon") 
plots_simp_n <- diversity(plots_cov_n,index="simpson")	#simpson diversity
plots_rich_n <- specnumber(plots_cov_n)					#species richness

plots_sh_n
plots_simp_n
plots_rich_n
```

**2023**

```{r}
plots_sh_n23 <- diversity(plots_covn_23,index="shannon") 
plots_simp_n23 <- diversity(plots_covn_23,index="simpson")	#simpson diversity
plots_rich_n23 <- specnumber(plots_covn_23)					#species richness

plots_sh_n23
plots_simp_n23
plots_rich_n23
```



### Exotic

**2019**

```{r}
sub_exotic_cover <- covdat %>% 
  select(any_of(names(exotic_wide)))
```

```{r}
p1_covex <- (2.5/7.5)*apply(sub_exotic_cover[1:10,],2,mean) + 
			(5/7.5)*apply(sub_exotic_cover[11:12,],2,mean)
	p2_covex <- (2.5/7.5)*apply(sub_exotic_cover[13:22,],2,mean) + 
			(5/7.5)*apply(sub_exotic_cover[23:24,],2,mean)
	p3_covex <- (2.5/7.5)*apply(sub_exotic_cover[25:34,],2,mean) + 
			(5/7.5)*apply(sub_exotic_cover[35:36,],2,mean)
	p4_covex <- (2.5/7.5)*apply(sub_exotic_cover[37:46,],2,mean) + 
			(5/7.5)*apply(sub_exotic_cover[47:48,],2,mean)
	
	plots_cov_ex <- rbind(p1_covex,p2_covex,p3_covex,p4_covex)
```

```{r}
plots_sh_ex <- diversity(plots_cov_ex,index="shannon") 	#shannon H'
plots_simp_ex <- diversity(plots_cov_ex,index="simpson")	#simpson diversity
plots_rich_ex <- specnumber(plots_cov_ex)					#species richness
	
plots_sh_ex
plots_simp_ex
plots_rich_ex
```


**2023**

```{r}
sub_exotic_cover23 <- covdat_2023 %>% 
  select(any_of(names(exotic_wide)))
```

```{r}
p1_cov_ex23 <- (2.5/7.5)*apply(sub_exotic_cover23[1:10,],2,mean) + 
			(5/7.5)*apply(sub_exotic_cover23[11:12,],2,mean)
	p2_cov_ex23 <- (2.5/7.5)*apply(sub_exotic_cover23[13:22,],2,mean) + 
			(2.5/5)*apply(sub_exotic_cover23[23,],2,mean)
	p3_cov_ex23 <- (2.5/7.5)*apply(sub_exotic_cover23[24:33,],2,mean) + 
			(2.5/5)*apply(sub_exotic_cover23[34,],2,mean)
	
		plots_cov_ex23 <- rbind(p1_cov_ex23,p2_cov_ex23,p3_cov_ex23)
```

```{r}
plots_sh_ex23 <- diversity(plots_cov_ex23,index="shannon") 	#shannon H'
plots_simp_ex23 <- diversity(plots_cov_ex23,index="simpson")	#simpson diversity
plots_rich_ex23 <- specnumber(plots_cov_ex23)					#species richness
	
plots_sh_ex23
plots_simp_ex23
plots_rich_ex23
```


**How does the rank order of diversity among plots vary depending on whether you use the Shannon or Simpson index? How does the pattern of native species diversity compare to exotic species diversity?**

Answer here. 






